---
title: 'Execise 2: Cadmium exceedance'
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Purpose of this exercise: try to solve this question:
*How many times did the Cadmium concentration in sediment exceed the safe limit (ERL; Lyons et al., 2017) in 2015?

***

<br>


### 2.1. Data: ICES


We will download some EMODnet Biology species observations with the Download Toolbox:

  * Go to <http://www.ices.dk/>
  * Click ‘Explore Marine Data’
  * Click ‘Environmental contaminants’
  * On line ‘Contaminants and biological effects’ click on ‘per parameter’
  * On search interface, search for 'Cd'
  * In total you should see 69000 measurements, but in different matrices.
  * So now select in ‘per matrix’ only sediment samples. Best is those ‘Sediment - <63 micron fraction’.
  * Click the 'download' button and save the file as "20190208_Cd_data_ICES.csv""


***

<br>


### 2.2. Data analysis in R

Now we can start analyzing these data in R


First, we load some R packages that we will use in this exercise
```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
library(dplyr)      # package for data manipulation
library(data.table) # package for fast reading of csv files
library(sf)         # simple features
library(mapview)    # interactive maps
library(ggplot2)    # plotting
```

read the data:
```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
cd_data <- read.csv("data/20190208_Cd_data_ICES.csv")

head(cd_data)

```

It seems that there are multiple units in this dataset!

```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
# it seems that there are different units used in this dataset!
# create table with unit occurrences
table(cd_data$Unit)

```

Ok, let's do a conversion by using some dplyr functions.

The `%>%` is the 'pipe' operator, and is used to combine multiple actions
This makes it easier to read:

x `%>%` y `%>%` z, means 

'take data x', do y with it and with this result, do z


```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
# manual conversion
cd_data <- cd_data %>%
  mutate(ugkg = case_when( Unit == 'ug/g'  ~ Value * 1000,
                           Unit == 'mg/kg' ~ Value * 1000,
                           Unit == 'g/g'   ~ Value * 1E9,
                           Unit == 'ug/kg' ~ Value
                           ),
         Exceeded = as.numeric(ugkg >= 1200)
  )

```

Make the data spatially aware (convert to 'sf' object):

```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
cd_data_sf <- st_as_sf(cd_data,
                       coords=c("Longitude..degrees_east.","Latitude..degrees_north."),
                       crs = 4326, # WGS84
                       remove = FALSE)


```

Plot the data:

```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
mapview(cd_data_sf, zcol = "Exceeded",
        viewer.suppress = FALSE)


```

Get back to our question:
What data did we need only the year 2015.
So convert to 'Date' format (Posixct)

```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}

# Now, what was the question? How many times is it exceeded in 2015?
cd_data_sf$Date_posixct <- as.POSIXct(cd_data_sf$dd.mm.yyyyThh.mm.ss,
                                             format = "%d/%m/%Y %H:%M",
                                             tz = "UTC")
```

A quick check of the data:
```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}

# quick check of the data
ggplot(cd_data_sf) +
  geom_histogram(aes(x = Date_posixct))

```

Subset the data to only include 2015: 
```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
cd_data_sf_2015 <- cd_data_sf %>%
  filter(Date_posixct > as.POSIXct("2015-01-01 01:00:00", tz="UTC") &
         Date_posixct < as.POSIXct("2016-01-01 01:00:00", tz="UTC"))

```

plot the data:
```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
# plotting the data
ggplot(cd_data_sf_2015) +
  geom_point(aes(x = Date_posixct, y = ugkg)) +
  geom_hline(yintercept =1200, color = "red")

```

some more interactive plots:
```{r message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
mapview(cd_data_sf_2015, zcol = "Exceeded",
        viewer.suppress = FALSE) # to view in RStudio

# where are the threshold values exceeded?
mapview(cd_data_sf_2015[cd_data_sf_2015$Exceeded ==1,], zcol = "Exceeded",
        viewer.suppress = FALSE) # to view in RStudio

```

So can you answer the question?

***

<br>


#### Exercise 1
How many times was the Cadmium threshold exceeded in 2015?


#### Advanced exercises

* In which estuaries are these located (look up on map)
# In which year are the most exceedances?
